Disclosing funding sources for open access publication fees: the Open APC

### 2016-06-04 article.Rmd
replaced CrossRef with Crossref after their rebranding

---
title: 'Disclosing funding sources for open access publication fees: the Open APC
  initiative'
author:
- Najko Jahn, Bielefeld University Library, Bielefeld University (TIB) - German National
  Library of Science and Technology, Hannover, Germany
- <a href="http://orcid.org/0000-0002-5111-2788">Marco Tullney</a>, Technische Informationsbibliothek
date: "7 April 2016"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    template: peerj.sty
  html_document:
    fig_caption: yes
    force_captions: yes
    number_sections: no
    theme: readable
bibliography: literatur.bib
csl: frontiers.csl
abstract: Publication fees in open access publishing hold a prominent place on the
  agenda of researchers, policy-makers, and academic publishers. This paper contributes
  to the evolving empirical basis on open access funding. It describes the Open APC
  initiative, in which German universities and research organizations share their
  expenditures for publication fees. As method, the initiative uses existing open
  data tools to aggregate and disseminate institutional spending on open access publication
  fees. In total, 29 German research organizations self-reported funding of 6,279
  open access journal articles, which amounted to 8,039,339 €. The average payment
  for each article was 1,280 €, and the median payment 1,209 €. Our data-set comprises
  only 53 articles in hybrid journals. With an indexing coverage of 99 %, the findings
  reveal that the DOI agency CrossRef provides both comprehensive bibliographic coverage
  of the funded open access journal literature and disambiguated names of journal
  titles and publishing houses. We show that authority control of these bibliographic
  information is particularly relevant for the comparative study of the economical
  effects of open access publishing.
---

@@

<!-- Publication fees, often paid by funders or universities, are a widely discussed business model for open access journals. Yet, how and to what extent these activities are effective in terms of the number of supported research articles and associated costs remains under debate. This study contributes to a growing empirical base about institutional spending on open access publication fees. More specifically, it uses self-reported cost data from German universities and research organisations to examine how much they spent, and how the so funded open access journal articles were distributed over publishers and journal titles using bibliographic information from CrossRef. Such an approach extends methods and improves data collection activities for researchers and practitioners alike, as well as contribute to a better understanding of factors affecting the analysis of publication fees in open access publishing. -->

@@

The aim of the study was to examine how much German universities and research organisations spent on open access publication fees until 2015. Using self-reported cost data from the Open APC initiative, the analysis focused on the amount of institutional payments for publication fees, and compared these expenditure data with data from related Austrian and UK initiatives, both in terms of size and proportion of the different journal types fully and hybrid open access journal. We also asked how thoroughly self-reported articles were indexed in CrossRef, a DOI minting agency for scholarly literature, and analysed how institutional spending per articles was distributed over publishers and journal titles.

@@

We fetched bibliographic metadata for each article from CrossRef on May 19, 2016, on the basis of the reported DOIs. Although the Open APC initiative gathered metadata representing publishers and journals from CrossRef as well, this information was retrieved at the time when the participating institutions submitted the cost information. The Open APC initiative tracked the date of data submission and the contributors for every data submission with Git, a version control system, increasingly used for enabling reproducible research [@Ram_2013], and made this information available via GitHub to be transparent over time. However, CrossRef regularly updated metadata to represent ongoing mergers of publishing houses or name changes. A prominent example in this regard was the merger of the two large publishing houses Springer Business + Media and Nature Publishing Group announced on May 6, 2015, that operated as Springer Nature at the time of our study. To reflect these dynamics in academic publishing, we decided to retrieve updated metadata from CrossRef for the whole Open APC data-set instead of re-using publisher and journal information contained in the Open APC data-set.

As a client, we used the R package rcrossref [@rcrossref], developed and maintained by the rOpenSci initiative[^6], to access CrossRef's REST API. We requested the XML-based format `application/vnd.crossref.unixsd+xml` in which full and abbreviated journal titles as well as the ISSN media types, the International Standard Serial Number used to identify journals, were distinguished. It also contained normalised publisher information, thus avoiding confusion about naming of publishing houses other studies were faced with when working with self-reported data [@woodward_2014]. In cases where no bibliographic information could be obtained, we used the Open APC values. Because CrossRef was not the only registration agency for DOIs, but also the agencies DataCite and Medra minted DOIs for scholarly work, we furthermore obtained the DOI agency for each article with the help of the rcrossref client.

@@

Data collection methods of the Open APC initiative and those of the others differed in some aspects. For instance, whereas the DOI was a mandatory element in the Open APC data template that the participating institution were required to report, publication identifier in the Wellcome Trust data was also added through automated compliance checks. Our first screening of the data-sets revealed that some articles lacked a DOI. For this reason, and as our main focus is institutional funding for publication fees in Germany, we decided only to compare German spending with that of other initiatives, but not its size and distribution over publisher and journal titles, as well as the indexing coverage in CrossRef.

@@

## CrossRef indexing{.unnumbered}

@@

agencies <- doi_miss %>% 
  group_by(doi_agency) %>% 
  summarise(n()) %>%
  rbind(c("CrossRef", nrow(cr)),.)

@@

To identify the sponsored open access journal articles, as well as to gather bibliographic metadata, DOIs were a mandatory part of the Open APC initiative's data collection activities. The participating institutions reported DOIs for `r my_apc %>% filter(!is.na(doi)) %>% nrow()`  out of `r nrow(my_apc)` articles. using these DOIs, we retrieved additional metadata from CrossRef for `r nrow(cr)` publications,  representing `r nrow(cr) / nrow(my_apc) * 100`  % of the total article volume. Articles for which no metadata could be obtained, were either registered with the DOI agency DataCite (10 articles) or Medra (two articles). For eight articles, our parser could not gather the XML resource, although these publications were registered with CrossRef. Seven DOIs reported to the Open APC initiative did not resolve.

@@

We used the DOI to automatically fetch publisher and journal names for each article from the CrossRef REST API. Table 4 shows the top ten publishers in terms of payments made that represent `r apc_publisher %>% filter(!publisher == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of the spending for publication fees. In total, payments were made to  `r my_apc %>% group_by(publisher) %>% summarize(n()) %>% nrow()` publishing houses. In comparison with data from the UK, full open access publishers have a greater share on total spending. @Pinfield_2015, for instance, reported remarkably lower numbers for the open access publishers MPDI AG, Copernicus GmbH, and Hindawi Publishing.

@@

Most of the funding of publication fees in Germany went to the publisher Springer Nature, most llikeky profiting from the merge with the open access publisher BioMed Central in 2008, and in 2015 between the well-established publishers Springer Science + Business Media and Nature Publishing Group. Using the CrossRef-Member-ID instead of the publisher name, we were able to differentiate between journals  formely published by Springer Science + Business Media and Nature Publishing Group. In terms of articles, the majority of payments made were for publications in journal that were formerly associated with Springer Science + Business Media. Springer Science + Business Media journals accounted for `r nrow(springer)`articles, representing `r nrow(springer) / nrow(springer_nat) * 100` of the total Springer Nature article volume in the spending data, and `r sum(springer$euro) / sum(springer_nat$euro) * 100` % of the amount spent. Median publication fee spending slightly differed between Springer Science + Business Media (`r median(springer$euro)` €) and Nature Publishing (`r median(nature$euro)` €). However, price variation was higher for Nature Publishing journals (SD = `r sd(nature$euro)`€) compared with that of the former Springer Science + Business Media titles (SD = `r sd(nature$euro)` €).

@@

In our study, CrossRef thoroughly indexed open access journal articles in the Open APC data-set. We could gather metadata representing publisher and journal titles for 99 % articles, and successfully merged these information with the Open APC cost data. Using metadata from CrossRef, therefore, reduces the extensive validation work of bibliographic information provided that the reporting of the DOI along with the expenses is made mandatory. Drawing on CrossRef would also increase the comparability of cost data for future negotiations with publishers on open access agreements, and the open access spending between open data initiatives that apply the same reporting standards, as its metadata represent the dynamic landscape of academic publishing in terms of ongoing mergers of publishing houses or name changes.

@@

<!-- The purpose of the Open APC initiative is to gain better insight on spending for open access. The main operational idea—to make data submission easy by building upon existing data and by using popular open tools and data bases—has proven to be a feasible approach and to scale well. Starting with a few institutions, the integration of further data contributors has not been a problem. While we use article metadata reported by publishers to CrossRef, we think it's best to collect information on how much has been spent at the sources: the institutions paying for APCs. We look forward to more data contributors to the initiative and to adapt this scheme to new analyses, additional data, or international sources. The Open APC initiative can serve as an example of the value that open data, open source software, and the knowledge and skills of tech-savvy librarians add to modern tasks of libraries.-->

@@

[^19]: We would like to thank Martin Fenner for pointing this out to us. See also the CrossRef API documentation <https://github.com/CrossRef/rest-api-doc/blob/master/rest_api.md>

>>>>

---
title: 'Disclosing funding sources for open access publication fees: the Open APC
  initiative'
author:
- Najko Jahn, Bielefeld University Library, Bielefeld University (TIB) - German National
  Library of Science and Technology, Hannover, Germany
- <a href="http://orcid.org/0000-0002-5111-2788">Marco Tullney</a>, Technische Informationsbibliothek
date: "7 April 2016"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: yes
    template: peerj.sty
  html_document:
    fig_caption: yes
    force_captions: yes
    number_sections: no
    theme: readable
bibliography: literatur.bib
csl: frontiers.csl
abstract: Publication fees in open access publishing hold a prominent place on the
  agenda of researchers, policy-makers, and academic publishers. This paper contributes
  to the evolving empirical basis on open access funding. It describes the Open APC
  initiative, in which German universities and research organizations share their
  expenditures for publication fees. As method, the initiative uses existing open
  data tools to aggregate and disseminate institutional spending on open access publication
  fees. In total, 29 German research organizations self-reported funding of 6,279
  open access journal articles, which amounted to 8,039,339 €. The average payment
  for each article was 1,280 €, and the median payment 1,209 €. Our data-set comprises
  only 53 articles in hybrid journals. With an indexing coverage of 99 %, the findings
  reveal that the DOI agency Crossref provides both comprehensive bibliographic coverage
  of the funded open access journal literature and disambiguated names of journal
  titles and publishing houses. We show that authority control of these bibliographic
  information is particularly relevant for the comparative study of the economical
  effects of open access publishing.
---

@@

<!-- Publication fees, often paid by funders or universities, are a widely discussed business model for open access journals. Yet, how and to what extent these activities are effective in terms of the number of supported research articles and associated costs remains under debate. This study contributes to a growing empirical base about institutional spending on open access publication fees. More specifically, it uses self-reported cost data from German universities and research organisations to examine how much they spent, and how the so funded open access journal articles were distributed over publishers and journal titles using bibliographic information from Crossref. Such an approach extends methods and improves data collection activities for researchers and practitioners alike, as well as contribute to a better understanding of factors affecting the analysis of publication fees in open access publishing. -->

@@

The aim of the study was to examine how much German universities and research organisations spent on open access publication fees until 2015. Using self-reported cost data from the Open APC initiative, the analysis focused on the amount of institutional payments for publication fees, and compared these expenditure data with data from related Austrian and UK initiatives, both in terms of size and proportion of the different journal types fully and hybrid open access journal. We also asked how thoroughly self-reported articles were indexed in Crossref, a DOI minting agency for scholarly literature, and analysed how institutional spending per articles was distributed over publishers and journal titles.

@@

We fetched bibliographic metadata for each article from Crossref on May 19, 2016, on the basis of the reported DOIs. Although the Open APC initiative gathered metadata representing publishers and journals from Crossref as well, this information was retrieved at the time when the participating institutions submitted the cost information. The Open APC initiative tracked the date of data submission and the contributors for every data submission with Git, a version control system, increasingly used for enabling reproducible research [@Ram_2013], and made this information available via GitHub to be transparent over time. However, Crossref regularly updated metadata to represent ongoing mergers of publishing houses or name changes. A prominent example in this regard was the merger of the two large publishing houses Springer Business + Media and Nature Publishing Group announced on May 6, 2015, that operated as Springer Nature at the time of our study. To reflect these dynamics in academic publishing, we decided to retrieve updated metadata from Crossref for the whole Open APC data-set instead of re-using publisher and journal information contained in the Open APC data-set.

As a client, we used the R package rcrossref [@rcrossref], developed and maintained by the rOpenSci initiative[^6], to access Crossref's REST API. We requested the XML-based format `application/vnd.crossref.unixsd+xml` in which full and abbreviated journal titles as well as the ISSN media types, the International Standard Serial Number used to identify journals, were distinguished. It also contained normalised publisher information, thus avoiding confusion about naming of publishing houses other studies were faced with when working with self-reported data [@woodward_2014]. In cases where no bibliographic information could be obtained, we used the Open APC values. Because Crossref was not the only registration agency for DOIs, but also the agencies DataCite and Medra minted DOIs for scholarly work, we furthermore obtained the DOI agency for each article with the help of the rcrossref client.

@@

Data collection methods of the Open APC initiative and those of the others differed in some aspects. For instance, whereas the DOI was a mandatory element in the Open APC data template that the participating institution were required to report, publication identifier in the Wellcome Trust data was also added through automated compliance checks. Our first screening of the data-sets revealed that some articles lacked a DOI. For this reason, and as our main focus is institutional funding for publication fees in Germany, we decided only to compare German spending with that of other initiatives, but not its size and distribution over publisher and journal titles, as well as the indexing coverage in Crossref.

@@

## Crossref indexing{.unnumbered}

@@

agencies <- doi_miss %>% 
  group_by(doi_agency) %>% 
  summarise(n()) %>%
  rbind(c("Crossref", nrow(cr)),.)

@@

To identify the sponsored open access journal articles, as well as to gather bibliographic metadata, DOIs were a mandatory part of the Open APC initiative's data collection activities. The participating institutions reported DOIs for `r my_apc %>% filter(!is.na(doi)) %>% nrow()`  out of `r nrow(my_apc)` articles. using these DOIs, we retrieved additional metadata from Crossref for `r nrow(cr)` publications,  representing `r nrow(cr) / nrow(my_apc) * 100`  % of the total article volume. Articles for which no metadata could be obtained, were either registered with the DOI agency DataCite (10 articles) or Medra (two articles). For eight articles, our parser could not gather the XML resource, although these publications were registered with Crossref. Seven DOIs reported to the Open APC initiative did not resolve.

@@

We used the DOI to automatically fetch publisher and journal names for each article from the Crossref REST API. Table 4 shows the top ten publishers in terms of payments made that represent `r apc_publisher %>% filter(!publisher == "Other") %>% nrow() / my_apc %>% nrow() * 100` % of the spending for publication fees. In total, payments were made to  `r my_apc %>% group_by(publisher) %>% summarize(n()) %>% nrow()` publishing houses. In comparison with data from the UK, full open access publishers have a greater share on total spending. @Pinfield_2015, for instance, reported remarkably lower numbers for the open access publishers MPDI AG, Copernicus GmbH, and Hindawi Publishing.

@@

Most of the funding of publication fees in Germany went to the publisher Springer Nature, most llikeky profiting from the merge with the open access publisher BioMed Central in 2008, and in 2015 between the well-established publishers Springer Science + Business Media and Nature Publishing Group. Using the Crossref-Member-ID instead of the publisher name, we were able to differentiate between journals  formely published by Springer Science + Business Media and Nature Publishing Group. In terms of articles, the majority of payments made were for publications in journal that were formerly associated with Springer Science + Business Media. Springer Science + Business Media journals accounted for `r nrow(springer)`articles, representing `r nrow(springer) / nrow(springer_nat) * 100` of the total Springer Nature article volume in the spending data, and `r sum(springer$euro) / sum(springer_nat$euro) * 100` % of the amount spent. Median publication fee spending slightly differed between Springer Science + Business Media (`r median(springer$euro)` €) and Nature Publishing (`r median(nature$euro)` €). However, price variation was higher for Nature Publishing journals (SD = `r sd(nature$euro)`€) compared with that of the former Springer Science + Business Media titles (SD = `r sd(nature$euro)` €).

@@

In our study, Crossref thoroughly indexed open access journal articles in the Open APC data-set. We could gather metadata representing publisher and journal titles for 99 % articles, and successfully merged these information with the Open APC cost data. Using metadata from Crossref, therefore, reduces the extensive validation work of bibliographic information provided that the reporting of the DOI along with the expenses is made mandatory. Drawing on Crossref would also increase the comparability of cost data for future negotiations with publishers on open access agreements, and the open access spending between open data initiatives that apply the same reporting standards, as its metadata represent the dynamic landscape of academic publishing in terms of ongoing mergers of publishing houses or name changes.

@@

<!-- The purpose of the Open APC initiative is to gain better insight on spending for open access. The main operational idea—to make data submission easy by building upon existing data and by using popular open tools and data bases—has proven to be a feasible approach and to scale well. Starting with a few institutions, the integration of further data contributors has not been a problem. While we use article metadata reported by publishers to Crossref, we think it's best to collect information on how much has been spent at the sources: the institutions paying for APCs. We look forward to more data contributors to the initiative and to adapt this scheme to new analyses, additional data, or international sources. The Open APC initiative can serve as an example of the value that open data, open source software, and the knowledge and skills of tech-savvy librarians add to modern tasks of libraries.-->

@@

[^19]: We would like to thank Martin Fenner for pointing this out to us. See also the Crossref API documentation <https://github.com/CrossRef/rest-api-doc/blob/master/rest_api.md>
